---
title: "rlefer"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{rlefer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

`rlefer` is a R package that provides an interface to the Jobard and Lefer (1997)
algorithm implemented in C++. You can use this algorithm to draw non-overlapping
and evenly-spaced curves in a flow field (or a "vector field" if you prefer to call it this way).

If you want deep details about how the algorithm works, there is
a scientific paper (Jobard and Lefer 1997) that describes it. But
you might find [this article useful as well](https://pedro-faria.netlify.app/posts/2024/2024-02-19-flow-even/en/index.html).

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# A first example

In order to draw any curve with `rlefer`, we first need to generate
a flow field. One important detail, this flow field needs
to be a square (width equal to height).

## Generating a flow field

As a first example, let's draw evenly-spaced curves in a flow field
with 120x120 dimensions. We can generate such flow field
with the `ambient` R package like this:

```{r}
library(ambient)
flow_field_width <- 120
set.seed(50)
flow_field <- noise_perlin(c(flow_field_width, flow_field_width))
```

To make the flow field stronger and more violent, I scale the values in the
flow field by multiplying them by $2\pi$:

```{r}
flow_field <- 2 * pi * flow_field
```

We can visualize this flow field as a grid of angle values, like this:

```{r}
library(tidyverse)

build_grid_df <- function(angles, n) {
  tibble(
    x = rep(seq_len(n), each = n),
    y = rep(seq_len(n), times = n),
    value = angles |> as.vector()
  )
}

visualize_grid <- function(grid, n){
  # Calculate the n^2 lines
  grid <- grid %>% 
    mutate(
      line_id = seq_len(nrow(grid)),
      xend = cos(value),
      yend = sin(value),
    )
  
  # Spread the lines across the grid
  grid <- grid %>% 
    mutate(
      xend = xend + x,
      yend = yend + y
    )
  
  # Plot these lines
  u <- "inches"
  a <- arrow(length = unit(0.025, u))
  ggplot(grid) +
    geom_segment(
      aes(
        x = x, y = y,
        xend = xend,yend = yend,
        group = line_id
      ),
      arrow = a
    ) +
    coord_cartesian(
      xlim = c(0,n), ylim = c(0,n)
    ) +
    theme_void()
}

grid <- build_grid_df(flow_field, flow_field_width)
visualize_grid(grid, flow_field_width)
```


# How curves are drawn in `rlefer`?

The functions from `rlefer` draws curves by walking through the flow field.
In other words, we start somewhere in the flow field, and
we start to walk by following the direction of the
angle values we encounter trough the flow field.


As we walk trough, we record the x and y coordinates of each step we
take, and, at the end of the process, we will have a sequence of
points, that we can connect to form a curve.
So each curve drawn by `rlefer` is essentially a sequence
of points that we can connect to visualize the curve
they represent.

If we do that with `ggplot2` for example, we can use geoms
such as `geom_path()` or `geom_line()` to visualize these
curves.


# Drawing evenly-spaced curves with `rlefer`

Now that we have our flow field prepared, we can start to
draw curves into this flow field.

Here, you want to decide if you want to draw curves that are not only
non-overlapping, but also, curves that are evenly-spaced between each
other.

If you want to avoid overlapping curves, and want to have full control over
the starting points of each curve, you should use the `rlefer::non_overlapping_curves()` function. This function uses
just the part of the Jobard and Lefer (1997) algorithm that
is responsible for the overlap checking to draw non-overlapping curves.

In contrast, if want to both avoid overlapping curves, and also want
your curves to be evenly-spaced between each other, the,
you should use the `rlefer::even_spaced_curves()` function, which
takes full advantage of the Jobard and Lefer (1997) algorithm
to draw curves that are both non-overlapping, and also,
are evenly-spaced between each other.

As our first example, let's consider we want to draw both
non-overlapping and evenly-spaced curves. That is why we
are going to use the `rlefer::even_spaced_curves()` function.

To use this function we want to set some variables. First,
we want to decide the point where in the flow field we
are going to start drawing curves. This a single point,
and going to be the starting point of the first point
drawn in the flow field.

So, if you ask the function to draw 100 curves, it
will first draw a single curve from the x and y
coordinates you gave as the starting point. And then,
the function will start to draw the remaining 99 curves,
by picking starting points that are `d_sep` of distance
from the curve. So all remaining curves are
automatically derived from this first curve drawn.

In our example, I will set the starting point to x = 45
and y = 24. These numbers need to be inside the flow field
bounds. So both `x < flow_field_width` and `y < flow_field_width`
needs to be true.

Let's start by drawing 100 curves. We will walk 40 steps (`n_steps`) in each curve, and in each step, and we will
walk a distance of 1% of the flow field width (`step_length`) in each step.
If you raise the number of steps taken, the curves drawn will be longer, otherwise,
they will look shorter.

Also, if you raise the distance taken in each step, your curve will
be longer. But the curve you get as result, might be not very precise, and
will probably look a bit ugly. If you set this number too low, you curve will look
short even if the number of steps taken for each curve are very high.
So, be careful with this number, don't set it
to high or too low. 1% of the flow field width is usually a good number.

Now, the distance between each curve (`d_sep`), which is called as the "separating
distance" here, will be `1.0`. The algorithm will constantly check if each curve
is at a minimum distance of `d_sep` from each of it's neighbors.


```{r setup}
library(rlefer)
library(ggplot2)

n_steps <- 40
n_curves <- 100
min_allowed_steps <- 5
step_length <- 0.01 * flow_field_width
d_sep <- 1.0

# The coordinates x = 45 and y = 24 are used as the starting point:
curves <- even_spaced_curves(
  45, 24,
  n_curves,
  n_steps,
  min_allowed_steps,
  step_length,
  d_sep,
  flow_field
)


ggplot(curves) +
  geom_path(
    aes(x, y, group = curve_id)
  ) +
  coord_cartesian(
    xlim = c(0, flow_field_width),
    ylim = c(0, flow_field_width)
  ) +
  theme_void()
```


If we raise the number of steps, we will get longer curves, and the
spatial distribution of the curves might change also. Because with
longer curves, new spots, or new positions become available
to the algorithm as candidates for starting points of new curves.

Also, if we raise the number of curves, then, there will be more curves
obviously in the image, and, as a result, the image will look more full.

For example, let's raise the number of steps to 70, and the number of
curves to 300:

```{r}
n_curves <- 300
n_steps <- 70

curves <- even_spaced_curves(
  45, 24,
  n_curves,
  n_steps,
  min_allowed_steps,
  step_length,
  d_sep,
  flow_field
)

ggplot(curves) +
  geom_path(
    aes(x, y, group = curve_id)
  ) +
  coord_cartesian(
    xlim = c(0, flow_field_width),
    ylim = c(0, flow_field_width)
  ) +
  theme_void()
```


We can continue to increase the number of curves to fill more of the image,
until it gets completely filled with curves.


# Drawing non-overlapping curves

You might get frustrated with `rlefer::even_spaced_curve()`,
because it selects the starting points for all derived
curves automatically for you. And you might want to have better control
over where each curve starts.

If that is your case,
then, you should use the `rlefer::non_overlapping_curves()`
function instead. This function accepts a list with starting points
for each curve.

In other words, each element of this input list, is another list
with two named elements (`x` and `y`) that contains the x and y
coordinates of a starting point. The number of curves you
want to draw is automatically derived from the length
of this input list.

So the function will draw a curve from each starting point
described in this list, and it will constantly check if
each curve is overlapping or not it's neighbors curves.

In the example below, I am using `runif()` to randomly select
starting points across the flow field.

```{r}
set.seed(80)
xs <- runif(n_curves)
set.seed(90)
ys <- runif(n_curves)
xs <- xs * flow_field_width
ys <- ys * flow_field_width
starting_points <- list()
for(i in seq_len(n_curves)) {
  starting_points[[i]] <- list(x = xs[i], y = ys[i])
}

curves <- non_overlapping_curves(
  starting_points,
  n_steps,
  5,
  step_length,
  d_sep,
  flow_field
)

ggplot(curves) +
  geom_path(
    aes(x, y, group = curve_id)
  ) +
  coord_cartesian(
    xlim = c(0, flow_field_width),
    ylim = c(0, flow_field_width)
  ) +
  theme_void()
```



# References

Jobard, Bruno, and Wilfrid Lefer. 1997. “Creating Evenly-Spaced Streamlines of Arbitrary Density.” In Visualization in Scientific Computing ’97, edited by Wilfrid Lefer and Michel Grave, 43–55. Vienna: Springer Vienna.

